Reparametrization of COM-Poisson Regression Models with Applications in the Analysis of Experimental Count Data

### 2018-10-30 presentations/2018botucatu/slides-mca2018.Rnw
Add slides for the event in Botucatu

\date{\small \today \\[0.2cm]
  \url{jreduardo@usp.br} | \url{edujrrib@gmail.com}}

@@

@@

\begin{frame}{Count data}

  Number of times an event occurs in the observation unit.\\[0.3cm]

  Random variables that assume non-negative integer values.\\[0.3cm]

  Let $Y$ be a counting random variable, so that $y=0,1,2,\ldots$\\[0.5cm]

  Examples in experimental researches:

  \begin{itemize}
    \item number of grains produced by a plant;
    \item number of fruits produced by a tree;
    \item number of insects on a particular cell;
    \item others.
  \end{itemize}

@@

<<approx-plot, fig.height=3, fig.width=6, out.width=".9\\textwidth", fig.cap="Quadratic errors for the approximation of the (a) expectation and (b) variance. Dotted lines represent the restriction for suitable approximations given by \\cite{Shmueli2005}.">>=
include_graphics("figure/approx-plot-1.pdf")
@

@@

\section{Case studies}

\begin{frame}[t]{Motivating data sets and data analysis}

  \begin{itemize}
    \item Three illustrative examples of count data analysis are
      reported.
      \begin{itemize} \normalsize
      \item Assessing toxicity of nitrofen in aquatic systems, an
        equidispersed example;
      \item Soil moisture and potassium doses on soybean culture, an
        overdispersed example; and
      \item Artificial defoliation in cotton phenology, an
        underdispersed example.
      \end{itemize}
    \item In the data analysis, we consider the COM-Poisson model in the
      two forms (original and new parametrization) and the quasi-Poisson
      regression model as alternative models for the standard Poisson
      regression model.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Cotton bolls data}

@@

\begin{frame}{Soybean data}

  \begin{columns}
    \begin{column}{0.4\textwidth}
      \begin{figure}
        \includegraphics[width=4cm]{./images/soybean}
      \end{figure}
    \end{column}
    \begin{column}{0.6\textwidth}
      {\bf Aim:} evaluate the effects of potassium doses applied
      to soil in different soil moisture levels;\\[0.2cm]
      {\bf Design:} factorial 5 $\times$ 3 in a randomized complete
      block design (5 blocks);\\[0.2cm]
      {\bf Experimental unit:} a pot with a plant;\\[0.2cm]
      {\bf Factors:}
      \begin{itemize} \small
        \item Potassium fertilization dose (\texttt{K}):
        \item Soil moisture level (\texttt{umid}):
      \end{itemize}\vspace{0.1cm}
      {\bf Response variable:} Total number of bean seeds per pot;\\
    \end{column}
  \end{columns}

@@

\begin{frame}{Nitrofen data}

  \begin{columns}
    \begin{column}{0.4\textwidth}
      \begin{figure}
        \includegraphics[width=4cm]{./images/zooplankton}
      \end{figure}
    \end{column}
    \begin{column}{0.6\textwidth}
      {\bf Aim:} measure the reproductive toxicity of the herbicide
      nitrofen on a species of zooplankton
      (\textit{Ceriodaphnia dubia});\\[0.2cm]
      {\bf Design:} completely randomized design, with 10
      replicates;\\[0.2cm]
      {\bf Experimental unit:} zooplankton animal;\\[0.2cm]
      {\bf Factors:}
      \begin{itemize} \small
        \item herbicide nitrofen dose (\texttt{dose});
      \end{itemize}\vspace{0.1cm}
      {\bf Response variable:} Total number of live offspring;\\
    \end{column}
  \end{columns}

@@

  \vspace{-0.4cm}
\begin{table}[!ht]
\centering \footnotesize
\caption{Parameter estimates (Est) and ratio between estimate and
  standard error (SE).}
\vspace{-0.1cm}
\label{tab:coef-ovos}
\begin{tabular}{lrrrrrrrr}
  \toprule
  & \multicolumn{2}{c}{Poisson} &
    \multicolumn{2}{c}{COM-Poisson} &
    \multicolumn{2}{c}{COM-Poisson$_\mu$} &
    \multicolumn{2}{c}{Quasi-Poisson} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Wed Jun 27 09:38:32 2018
 & Est & Est/SE & Est & Est/SE & Est & Est/SE & Est & Est/SE \\
  \midrule
$\beta_{0}$ & 3.477 & 62.817 & 3.649 & 4.850 & 3.477 & 64.308 & 3.477 & 61.860 \\
  $\beta_{1}$ & $-$0.086 & $-$0.433 & $-$0.091 & $-$0.448 & $-$0.088 & $-$0.452 & $-$0.086 & $-$0.426 \\
  $\beta_{2}$ & 0.153 & 0.863 & 0.161 & 0.878 & 0.155 & 0.894 & 0.153 & 0.850 \\
  $\beta_{3}$ & $-$0.097 & $-$2.398 & $-$0.102 & $-$2.229 & $-$0.098 & $-$2.464 & $-$0.097 & $-$2.361 \\
   \bottomrule
\end{tabular}
\end{table}
  \vspace{-0.4cm}

<<pred-ovos, fig.height=3.5, fig.width=5.5, out.width="0.8\\textwidth", fig.cap="Number of live offsprings observed for each nitrofen concentration level with fitted curves and 95\\% confidence intervals.">>=
include_graphics("figure/pred-ovos-1")
@

@@

\subsection*{}

\begin{frame}{Additional results}

@@

<<comp-times, fig.height=4, fig.width=8, fig.cap="Computational times to fit the models under original and reparametrized versions based on the fifty repetitions.">>=

library(latticeExtra)
source("../codes/lattice-panels.R")

benchs <- readRDS("../codes/comparetimes.rds")
xlabs <- expression("COM-Poisson", "COM-Poisson"[mu])
bwplot(time ~ expr | case,
       ylab = "Time (seconds)",
       scales = list(
           y = list(relation = "free"),
           x = list(at = 1:2, labels = xlabs)
       ),
       data = benchs)

@@

\section*{Bibliography}
\begin{frame}[allowframebreaks]{References}
  \small
  \vspace*{-0.5cm}
  \bibliography{../references.bib}
\end{frame}

>>>>

\date{\small 11th October, 2018 - UNESP/Botucatu \\[0.2cm]
  \url{clarice.demetrio@usp.br} |\url{jreduardo@usp.br}}

@@

\section{Motivating data sets}

%-------------------------------------------
\begin{frame}{Artificial defoliation in cotton phenology}

\begin{frame}{Artificial defoliation in cotton phenology}
\vspace{-0.2cm}

<<desc-cotton, fig.width=9, fig.height=5.8, fig.cap="(a) Perfis das contagens de ninfas observadas para cada bloco em cada uma das quatro cultivares BRS, as linhas em preto representam os perfis médios. (b) Dispersão das médias (eixo $x$) e variâncias (eixo $y$) amostrais, na escala logarítmica, calculadas sobre os blocos em cada tempo e cultivar para cada variável de interesse, a linha pontilhada representa a reta média $=$ variância (equidispersão) e a contínua representa um ajuste de mínimos quadrados.">>=
include_graphics("figure/desc-cotton-1.pdf")
@
\end{frame}

%-------------------------------------------
\begin{frame}{Potassium fertilization on soybean culture}
  \begin{columns}
    \begin{column}{0.4\textwidth}
      \begin{figure}
        \includegraphics[width=4cm]{./images/soybean}
      \end{figure}
    \end{column}
    \begin{column}{0.6\textwidth}
      {\bf Aim:} evaluate the effects of potassium doses applied
      to soil in different soil moisture levels;\\[0.2cm]
      {\bf Design:} factorial 5 $\times$ 3 in a randomized complete
      block design (5 blocks);\\[0.2cm]
      {\bf Experimental unit:} a pot with a plant;\\[0.2cm]
      {\bf Factors:}
      \begin{itemize} \small
        \item Potassium fertilization dose (\texttt{K}):
        \item Soil moisture level (\texttt{umid}):
      \end{itemize}\vspace{0.1cm}
      {\bf Response variable:} Total number of bean seeds per pot;\\
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Potassium fertilization on soybean culture}
\vspace{-0.3cm}
<<desc-soya, fig.width=9, fig.height=6, fig.cap="(a) Number of bean seeds per pot for each potassium dose and moisture level (left) and (b) sample mean against sample variance of the five replicates for each experimental treatment. Solid lines are smoothing curves and the least-squares line.">>=
include_graphics("figure/desc-soy-1.pdf")
@
\end{frame}

%-------------------------------------------
\begin{frame}{Assessing toxicity of nitrofen in aquatic systems}
  \begin{columns}
    \begin{column}{0.4\textwidth}
      \begin{figure}
        \includegraphics[width=4cm]{./images/zooplankton}
      \end{figure}
    \end{column}
    \begin{column}{0.6\textwidth}
      {\bf Aim:} measure the reproductive toxicity of the herbicide
      nitrofen on a species of zooplankton
      (\textit{Ceriodaphnia dubia});\\[0.2cm]
      {\bf Design:} completely randomized design, with 10
      replicates;\\[0.2cm]
      {\bf Experimental unit:} zooplankton animal;\\[0.2cm]
      {\bf Factors:}
      \begin{itemize} \small
        \item herbicide nitrofen dose (\texttt{dose});
      \end{itemize}\vspace{0.1cm}
      {\bf Response variable:} Total number of live offspring;\\
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Assessing toxicity of nitrofen in aquatic systems}
<<desc-nitrofen, results="asis", fig.width=8, fig.height=3.5, fig.cap="(a) Number of live offsprings observed for each nitrofen concentration level and (b) scatterplot of the sample means against sample variances.">>=
include_graphics("figure/desc-nitrofen-1")
@
\end{frame}

@@

@@

<<approx-plot, fig.height=3, fig.width=6, out.width=".9\\textwidth", fig.cap="Errors for the approximation of the (a) expectation and (b) variance. Dotted lines represent the restriction for suitable approximations given by \\cite{Shmueli2005}.">>=
include_graphics("figure/approx-plot-1.pdf")
@

@@

\section{Data analysis results}

@@

@@

@@

@@

  \vspace{-0.4cm}
\begin{table}[!ht]
\centering \scriptsize
\caption{Parameter estimates (Est) and ratio between estimate and
  standard error (SE).}
\vspace{-0.1cm}
\label{tab:coef-ovos}
\begin{tabular}{lrrrrrrrr}
  \toprule
  & \multicolumn{2}{c}{Poisson} &
    \multicolumn{2}{c}{COM-Poisson} &
    \multicolumn{2}{c}{COM-Poisson$_\mu$} &
    \multicolumn{2}{c}{Quasi-Poisson} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Wed Jun 27 09:38:32 2018
 & Est & Est/SE & Est & Est/SE & Est & Est/SE & Est & Est/SE \\
  \midrule
$\beta_{0}$ & 3.477 & 62.817 & 3.649 & 4.850 & 3.477 & 64.308 & 3.477 & 61.860 \\
  $\beta_{1}$ & $-$0.086 & $-$0.433 & $-$0.091 & $-$0.448 & $-$0.088 & $-$0.452 & $-$0.086 & $-$0.426 \\
  $\beta_{2}$ & 0.153 & 0.863 & 0.161 & 0.878 & 0.155 & 0.894 & 0.153 & 0.850 \\
  $\beta_{3}$ & $-$0.097 & $-$2.398 & $-$0.102 & $-$2.229 & $-$0.098 & $-$2.464 & $-$0.097 & $-$2.361 \\
   \bottomrule
\end{tabular}
\end{table}
  \vspace{-0.3cm}

<<pred-ovos, fig.height=3.5, fig.width=5.5, out.width="0.65\\textwidth", fig.cap="Number of live offsprings observed for each nitrofen concentration level with fitted curves and 95\\% confidence intervals.">>=
include_graphics("figure/pred-ovos-1")
@

@@

\subsection{Additional results}

\begin{frame}{Computational times for fitting}

@@

<<comp-times, fig.height=4, fig.width=8, fig.cap="Computational times to fit the models under original and reparametrized versions based on the fifty repetitions.">>=
include_graphics("figure/comp-times-1")
@

@@

\section*{Bibliography}
\begin{frame}[allowframebreaks]{References}
  \small
  \vspace*{-0.5cm}
  \bibliography{../../references.bib}
\end{frame}
